Online Verification of Commutativity

### 2020-08-11 acmPaper/paper.tex
Move appendix to appendix
fix figures:

\labelformat{algocf}{Algorithm~#1}

@@

\section{Solving the Online Addition Problem}

We present a polynomial time solution to the \textsc{verification set} problem.
\xxx[as]{Elsewhere, ``problem'' is also in small-caps. Either way is fine, but consistency is important.}
Like in the online baseline algorithm, we do not concern ourselves with parallel pairs where neither or both paths pass through the new edge.
The key observation that allows us to improve on the online baseline is that, for a given source and sink pair, only a single parallel pair needs to be verified.
This is an implication of Theorem~\ref{reductionRule}, expanded on later.
Theorem~\ref{verifyingSet} shows that should our selected set of pairs along with cycles passing through the new edge be verified commutative, the entire diagram must commute.
The approach is to identify a parallel pair with exactly one path through the edge for each (source, sink) pair (\ref{algo_online_polynomial}).
\xxx[as]{Instead of putting the reference in parens, say ``Algorithm X is the strategy for doing Y. Some more details include Z.''}

@@

The try block is executed at most $O(|V|^2)$ times, so that this is the bound on the number of pairs verified.
%
The bound is asymptotically tight. This can be seen in the case where the graph contains $2N$ nodes besides $S$ and $T$.
\xxx[as]{Introduce this concept by saying that ``we imagine dividing the nodes into two groups...''}
We consider $N$ of the nodes to be in group 1 and the other $N$ to be in group 2. Every node in group 1 has a forward node to every node in group 2 as well as to $S$. $T$ has a forward edge to every node in group 2. In this diagram, on the addition of edge $(S, T)$, $N^2$ paths need to be verified which is polynomial in the total number of nodes, $2N+2$.

Also notice that if trying to optimize for path length (say, if composing functions is expensive) then ``find any path'' can be replaced with ``find the shortest path.''

An efficient implementation of the algorithm can run in $O(|V|^2(|V|+|E|))$ time, with space complexity not exceeding the asymptotic $O(|V|^2)$ bound on the output.
In such an implementation, path finding from a given source node to all potential sink nodes could be done in a single $O(|V|+|E|)$ breadth first search.

@@

We observe that there are some redundancies in the diagram.
%
Consider the situation in Figure~\ref{figure_reduction_rule}.
\xxx[as]{This needs a bit more explanation. What is the reader looking for? Also, the caption of this figure could probably be a little clearer.}

\begin{figure}
\begin{center}
\begin{tikzpicture}[scale=0.2]
\tikzstyle{every node}+=[inner sep=0pt]
\draw [black] (16.6,-34.1) circle (3);
\draw (16.6,-34.1) node {$S$};
\draw [black] (56.8,-33.6) circle (3);
\draw (56.8,-33.6) node {$T$};
\draw [black] (28,-20.1) circle (3);
\draw (28,-20.1) node {$P_1$};
\draw [black] (44.5,-20.1) circle (3);
\draw (44.5,-20.1) node {$P_2$};
\draw [black] (28,-44.9) circle (3);
\draw (28,-44.9) node {$Q_1$};
\draw [black] (44.5,-44.9) circle (3);
\draw (44.5,-44.9) node {$Q_2$};
\draw [black] (31,-20.1) -- (41.5,-20.1);
\fill [black] (41.5,-20.1) -- (40.7,-19.6) -- (40.7,-20.6);
\draw (36.25,-20.6) node [below] {$g_1$};
\draw [black] (26.11,-22.43) -- (18.49,-31.77);
\fill [black] (18.49,-31.77) -- (19.39,-31.47) -- (18.61,-30.84);
\draw (21.74,-25.67) node [left] {$f_1$};
\draw [black] (25.82,-42.84) -- (18.78,-36.16);
\fill [black] (18.78,-36.16) -- (19.01,-37.08) -- (19.7,-36.35);
\draw (23.57,-39.02) node [above] {$f_2$};
\draw [black] (31,-44.9) -- (41.5,-44.9);
\fill [black] (41.5,-44.9) -- (40.7,-44.4) -- (40.7,-45.4);
\draw (36.25,-45.4) node [below] {$g_2$};
\draw [black] (54.59,-35.63) -- (46.71,-42.87);
\fill [black] (46.71,-42.87) -- (47.64,-42.7) -- (46.96,-41.96);
\draw (49.21,-38.76) node [above] {$h_2$};
\draw [black] (54.78,-31.38) -- (46.52,-22.32);
\fill [black] (46.52,-22.32) -- (46.69,-23.25) -- (47.43,-22.57);
\draw (51.19,-25.39) node [right] {$h_1$};
\draw [black] (28,-23.1) -- (28,-41.9);
\fill [black] (28,-41.9) -- (28.5,-41.1) -- (27.5,-41.1);
\draw (27.5,-32.5) node [left] {$l$};
\draw [black] (19.6,-34.06) -- (53.8,-33.64);
\fill [black] (53.8,-33.64) -- (52.99,-33.15) -- (53.01,-34.15);
\draw [black] (44.5,-41.9) -- (44.5,-23.1);
\fill [black] (44.5,-23.1) -- (44,-23.9) -- (45,-23.9);
\draw (45,-32.5) node [right] {$m$};
\end{tikzpicture}
Each arrow represents a path, and $(S,T)$ is the new edge being added.
\end{center}
\caption{Reduction rule.}
\label{figure_reduction_rule}
\end{figure}

@@

The proof \xxx[as]{which proof?} is not affected if any of these \xxx[as]{which?} paths is the identity, e.g., if $f_1$ is the identity and $S$ and $P_1$ are actually the same node.
\xxx[as]{Maybe this observation belongs somewhere else, where we are talking about the correctness argument/proof?}

@@

Under the assumption that the only path operations allowed are composition and replacement of one path by a different, equal path, as would be true when edges are generic functions, and no other information is available, so that $F$ is a semi-group, this ``reduction'' rule is also the only rule to reduce the set of path pairs to check by finding implications.

@@

Using this information it is possible to choose a minimal subset of path pairs to verify, as in \ref{algo_online_minimal}.

We construct a graph with a node for each possible (source, sink) pair in the graph- each node then represents a possible choice for parallel pair endpoint pairs, and greedily search for the smallest set of nodes from which the entire graph would be reachable. The idea is to look for ``roots'' in the graph that have to be included in the ultimate verification set because they have no predecessor an cannot be verified ``through'' the verification of some other pair. Then all the successors whose verification is implied by the roots are eliminated.

@@

\begin{figure}
    % TODO format nicely.
    \begin{subfigure}{0.5\linewidth}
      \includegraphics[width=.8\linewidth]{timeVsSize_10_OptimalSet.png}
      \caption{~\ref{algo_online_minimal}}
      \label{fig:sfigOptimalTvsS}
    \end{subfigure}

    \begin{subfigure}{0.5\linewidth}
      \includegraphics[width=.8\linewidth]{timeVsSize_10_Polynomial.png}
      \caption{~\ref{algo_online_polynomial}}
      \label{fig:sfigPolynomialTvsS}
    \end{subfigure}

    \begin{subfigure}{0.5\linewidth}
      \includegraphics[width=.8\linewidth]{timeVsSize_10_NaiveChecker.png}
      \caption{Naive baseline}
      \label{fig:sfigNa\"{i}veTvsS}
    \end{subfigure}

    \begin{subfigure}{0.5\linewidth}
      \includegraphics[width=.8\linewidth]{timeVsSize_10_TwoFlipPathChecker.png}
      \caption{Two flip tolerant baseline}
      \label{fig:sfigTwoFlipTvsS}
    \end{subfigure}

    \begin{subfigure}{0.5\linewidth}
      \includegraphics[width=.8\linewidth]{timeVsSize_10_BatchChecker.png}
      \caption{Batch algorithm baseline}
      \label{fig:sfigBatchTvsS}
    \end{subfigure}
    \caption{Plots of scaling of time with input size and density}
    \label{fig:timeVsSize}
\end{figure}

\subsection{Variance}
The periodic spikes in Figure ~\ref{fig:sfigOptimalTvsS} are striking. 
We plot the spread of results to understand what is happening.
We find ~\ref{algo_online_minimal} has outliers about two standard deviation above the mean responsible for the spikes in the average.
The outliers themselves follow a polynomial curve.
The plots in Figure ~\ref{fig:variance} depict the situation.
Grey points are the results of evaluation on individual points, and error bars show standard deviation. 
The black curve traces the mean.
No such effect is observed in ~\ref{algo_online_polynomial}.
\begin{figure}
    % TODO format nicely.
    \begin{subfigure}{0.5\linewidth}
      \includegraphics[width=.8\linewidth]{variance_10_OptimalSet.png}
      \caption{~\ref{algo_online_minimal}}
      \label{fig:sfigOptimalSpread}
    \end{subfigure}

    \begin{subfigure}{0.5\linewidth}
      \includegraphics[width=.8\linewidth]{variance_10_Polynomial.png}
      \caption{~\ref{algo_online_polynomial}}
      \label{fig:sfigPolynomialSpread}
    \end{subfigure}

    \begin{subfigure}{0.5\linewidth}
      \includegraphics[width=.8\linewidth]{variance_10_TwoFlip.png}
      \caption{Two flip tolerant baseline}
      \label{fig:sfigTwoFlipSpread}
    \end{subfigure}

@@

\xxx{Absolutely nothing related to this was ever done.}

>>>>

\labelformat{algocf}{algorithm\,(#1)}

@@

\section{Solving the online addition problem}

We present a polynomial time solution to the \textsc{verification set} problem.
Like with the baseline, we do not concern ourselves with parallel pairs where neither or both paths pass through the new edge.
The key observation that allows us to improve on the online baseline is that, for a given source and sink pair, only a single parallel pair needs to be verified. 
This is an implication of Theorem \ref{reductionRule}, expanded on later.
Theorem \ref{verifyingSet} shows that should our selected set of pairs along with cycles passing through the new edge be verified commutative, the entire diagram must commute. 
The approach is to identify a parallel pair with exactly one path through the edge for each (source, sink) pair (algorithm ~\ref{algo_online_polynomial}).

@@

The try block is executed at most $O(|V|^2)$ times, so that this is the bound on the number of pairs verified.

The bound is asymptotically tight. This can be seen in the case where the graph contains 2N nodes besides S and T. We consider N of the nodes to be in group 1, and the other N to be in group 2. Every node in group 1 has a forward node to every node in group 2, as well as to S. T has a forward edge to every node in group 2. In this diagram, on the addition of edge (S, T), $N^2$ paths need to be verified which is polynomial in 2N+2.

Also notice that if trying to optimize for path length (say, if composing functions is expensive) then "find any path" can be replaced with "find shortest path".

An efficient implementation of the algorithm can run in $O(|V|^2(|V|+|E|))$ time, space complexity not exceeding the asymptotic $O(|V|^2)$ bound on the output.
In such an implementation, path finding from a given source node to all potential sink nodes could be done in a single $O(|V|+|E|)$ breadth first search.

@@

We observe that there are some redundancies in the diagram.

Consider the situation in Figure ~\ref{figure_reduction_rule}.

\begin{figure}
\begin{center}
\begin{tikzpicture}[scale=0.2]
\tikzstyle{every node}+=[inner sep=0pt]
\draw [black] (16.6,-34.1) circle (3);
\draw (16.6,-34.1) node {$S$};
\draw [black] (56.8,-33.6) circle (3);
\draw (56.8,-33.6) node {$T$};
\draw [black] (28,-20.1) circle (3);
\draw (28,-20.1) node {$P_1$};
\draw [black] (44.5,-20.1) circle (3);
\draw (44.5,-20.1) node {$P_2$};
\draw [black] (28,-44.9) circle (3);
\draw (28,-44.9) node {$Q_1$};
\draw [black] (44.5,-44.9) circle (3);
\draw (44.5,-44.9) node {$Q_2$};
\draw [black] (31,-20.1) -- (41.5,-20.1);
\fill [black] (41.5,-20.1) -- (40.7,-19.6) -- (40.7,-20.6);
\draw (36.25,-20.6) node [below] {$g_1$};
\draw [black] (26.11,-22.43) -- (18.49,-31.77);
\fill [black] (18.49,-31.77) -- (19.39,-31.47) -- (18.61,-30.84);
\draw (21.74,-25.67) node [left] {$f_1$};
\draw [black] (25.82,-42.84) -- (18.78,-36.16);
\fill [black] (18.78,-36.16) -- (19.01,-37.08) -- (19.7,-36.35);
\draw (23.57,-39.02) node [above] {$f_2$};
\draw [black] (31,-44.9) -- (41.5,-44.9);
\fill [black] (41.5,-44.9) -- (40.7,-44.4) -- (40.7,-45.4);
\draw (36.25,-45.4) node [below] {$g_2$};
\draw [black] (54.59,-35.63) -- (46.71,-42.87);
\fill [black] (46.71,-42.87) -- (47.64,-42.7) -- (46.96,-41.96);
\draw (49.21,-38.76) node [above] {$h_2$};
\draw [black] (54.78,-31.38) -- (46.52,-22.32);
\fill [black] (46.52,-22.32) -- (46.69,-23.25) -- (47.43,-22.57);
\draw (51.19,-25.39) node [right] {$h_1$};
\draw [black] (28,-23.1) -- (28,-41.9);
\fill [black] (28,-41.9) -- (28.5,-41.1) -- (27.5,-41.1);
\draw (27.5,-32.5) node [left] {$l$};
\draw [black] (19.6,-34.06) -- (53.8,-33.64);
\fill [black] (53.8,-33.64) -- (52.99,-33.15) -- (53.01,-34.15);
\draw [black] (44.5,-41.9) -- (44.5,-23.1);
\fill [black] (44.5,-23.1) -- (44,-23.9) -- (45,-23.9);
\draw (45,-32.5) node [right] {$m$};
\end{tikzpicture}
Each arrow represents a path, and (S,T) is the new edge being added.
\end{center}
\caption{Reduction rule.}
\label{figure_reduction_rule}
\end{figure}

@@

The proof is not affected if any of these paths is the identity, eg. if $f_1$ is the identity and S and $P_1$ are actually the same node.

@@

Under the assumption that the only path operations allowed are composition and replacement of one path by a different, equal path, as would be true when edges are generic functions, and no other information is available, so that F is a semi-group, this "reduction" rule is also the only rule to reduce the set of path pairs to check by finding implications.

@@

Using this information it is possible to choose a minimal subset of path pairs to verify, as in algorithm ~\ref{algo_online_minimal}.

We construct a graph with a node for each possible (source, sink) pair in the graph- each node then represents a possible choice for parallel pair endpoint pairs, and greedily search for the smallest set of nodes from which the entire graph would be reachable. The idea is to look for "roots" in the graph that have to be included in the ultimate verification set because they have no predecessor an cannot be verified "through" the verification of some other pair. Then all the successors whose verification is implied by the roots are eliminated.

@@

\begin{figure}
    % TODO format nicely.
    \begin{subfigure}{\linewidth}
      \includegraphics[width=\linewidth]{timeVsSize_10_OptimalSet.png}
      \caption{~\ref{algo_online_minimal}}
      \label{fig:sfigOptimalTvsS}
    \end{subfigure}

    \begin{subfigure}{\linewidth}
      \includegraphics[width=\linewidth]{timeVsSize_10_Polynomial.png}
      \caption{~\ref{algo_online_polynomial}}
      \label{fig:sfigPolynomialTvsS}
    \end{subfigure}

    \begin{subfigure}{\linewidth}
      \includegraphics[width=\linewidth]{timeVsSize_10_NaiveChecker.png}
      \caption{Naive baseline}
      \label{fig:sfigNa\"{i}veTvsS}
    \end{subfigure}

\end{figure}
\begin{figure}\ContinuedFloat

    \begin{subfigure}{\linewidth}
      \includegraphics[width=\linewidth]{timeVsSize_10_TwoFlipPathChecker.png}
      \caption{Two flip tolerant baseline}
      \label{fig:sfigTwoFlipTvsS}
    \end{subfigure}

    \begin{subfigure}{\linewidth}
      \includegraphics[width=\linewidth]{timeVsSize_10_BatchChecker.png}
      \caption{Batch algorithm baseline}
      \label{fig:sfigBatchTvsS}
    \end{subfigure}
    \caption{Plots of scaling of time with input size and density}
    \label{fig:timeVsSize}
\end{figure}

\subsection{Variance}
The periodic spikes in Figure ~\ref{fig:sfigOptimalTvsS} are striking. 
We plot the spread of results in Figure~\ref{fig:sfigOptimalSpread} to understand what is happening.
Grey points are the results of evaluation on individual points, and error bars show standard deviation. 
The black curve traces the mean.
We find ~\ref{algo_online_minimal} has outliers about two standard deviation above the mean responsible for the spikes in the average.
The outliers themselves follow a polynomial curve, appearing almost periodically.
We have not yet identified the cause of the behavior.
Figure ~\ref{fig:variance} depicts the situation for ~\ref{algo_online_polynomial}, where no such effect is observed.

\begin{figure}
    % TODO format nicely.
    \begin{subfigure}{\linewidth}
      \includegraphics[width=\linewidth]{variance_10_OptimalSet.png}
      \caption{~\ref{algo_online_minimal}}
      \label{fig:sfigOptimalSpread}
    \end{subfigure}

    \begin{subfigure}{\linewidth}
      \includegraphics[width=\linewidth]{variance_10_Polynomial.png}
      \caption{~\ref{algo_online_polynomial}}
      \label{fig:sfigPolynomialSpread}
    \end{subfigure}

    % \begin{subfigure}{\linewidth}
    %   \includegraphics[width=\linewidth]{variance_10_TwoFlip.png}
    %   \caption{Two flip tolerant baseline}
    %   \label{fig:sfigTwoFlipSpread}
    % \end{subfigure}

@@

As described in section ~\ref{sec:}
